Visualizing Project Evolution Through Abstract Syntax Tree Analysis

### 2016-04-09 paper/paper.tex
Add some figures to latex.
Update tex file.

% Copied from:
% http://mirror.its.dal.ca/ctan/macros/latex/contrib/IEEEtran/bare_conf.tex
\documentclass[conference]{IEEEtran}
% Some Computer Society conferences also require the compsoc mode option,
% but others use the standard conference format.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}
\usepackage{changepage}

@@

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation
% can be obtained at:
% http://www.ctan.org/pkg/graphicx
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found at:
% http://www.ctan.org/pkg/epslatex
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
Analysing the changes in software repositories over time is difficult since changes come from different developers and affect different parts of a repository.
Individual commits provide the changed files, code diffs and a message.
However, when many commits are viewed as a group, the details are lost.
Currently there is a lack of tools for tracking the changes in a repository beyond simply looking at the number of commits of lines changed.
The commit information can be misleading since lines of code gives no indication of what was actually being worked on without careful examination of the changed code.
Providing a more comprehensive view of this information could provide deeper insights into how software projects evolve since changes to design and features are not clearly visible.

We present a novel tool for visualizing Java projects.
It provides more detailed information regarding all of the changes along the master branch of a project.
It visualizes changes in type usage over time for projects and developers and derives detailed statistics using the differences between Abstract Syntax Trees (ASTs).
We are tracking additions and deletions of declarations and invocations for each type.
The AST diff between commits is a reentrant algorithm that can be incrementally calculated as new commits are made.

Using this tool we show the changes in type coverage and other usage statistics over time in several popular repositories.
Type coverage information from the AST is compared to file coverage to check for correlation or if there is unique information provided by type information.
Potential uses for such a tool for managers and developers is discussed.
\end{abstract}

@@

The history and evolution of software repositories are often tracked by version control systems. The information is generally kept in the form of commits, which show the differences between each subsequent version of the software.  These services become vital for large projects since individual changes would be impossible to track manually. Large projects have thousands of individual changes made by many different developers. With more changes to look through, it becomes more difficult to see how a project is evolving over time since there is so much information to look through. Most hosting services for git repositories provide some basic tools for analysing projects over time, but too much information is lost due to the lack of fine grained details. The information regarding contributions from developers are also misleading since the number of commits and lines of code do not necessarily reflect meaningful changes to a repository.

In Java, a type can be a primitive type or an object reference. Primitive types are those that are built into the language such as an int or char. Object references point to instances of classes in memory which are defined in the project or through an included library. The types in a project represent the functionality available for a developer to work with. A large or complex project may have many different types, whereas a small project might only use a few. Different parts of a program are going to use different types and libraries. When writing or contributing to a project, the author’s choice of which part of the program to edit will in turn affect what types they end up working with.

@@

Type information is useful for seeing what libraries and functions are being used.  When a part of code becomes deprecated, continuing to use it may lead to design or  security flaws. Viewing the type information of a change can show when deprecated classes are added or removed from a project as well as how many remain.

@@

Robert Dyer \textit{et al.} \cite{Dyer:2014:MBA:2568225.2568295} mined AST nodes to study the use of new Java language features over time. (More in challenge paper)

@@

Patrick Wagstrom \textit{et al.} \cite{Patrick:Wagstrom:2012} explored the roles that developers take in a networked, social development environments like GitHub. These role were defined as their level of contribution as well as the types of issues they dealt with. Developers will take on multiple roles in a project and will sometimes fulfill the same role across different projects

@@

@@

In order to gather all the changes for a given repository over time we generated AST for each commit and compared the difference. We did this by running ‘git log’ over the git repository and listing all the commit ids in increasing time. Each commit and its parent were then compared. If the commit had no parent then every file in the commit was treated as an addition. Since we compared each commit to its parent we were able to account for branching that happened in the master branch of the repository. Also we ignored commits that were merges as all changes in the merge should already be accounted for in previous commits.

@@

In the following sections, we present {N} case studies demonstrating the extra insight that TypeV makes available.

@@

We chose Antlr4 as it is a substantial Java project, with a commit history spanning several years. We can also see distinct changes throughout its history.

For example, in December of 2011 we can see that ParserATNSimulator has a large amount of additions and deletions in roughly equal proportion (Figure~\ref{fig:parser1}. This suggests that ParserATNSimulator is not being removed from the project but instead there is some refactoring being done. If we take a closer look at the commit messages we can see that they mention refactoring and reorganization of code (Figure~\ref{fig:parser2}). Zooming in on that date,  shows that during that month they were finishing work on a new version of ParserATNSimulator.

@@

@@

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
\nocite{*}
\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{bibliography,autogen}

>>>>

% Copied from:
% http://mirror.its.dal.ca/ctan/macros/latex/contrib/IEEEtran/bare_conf.tex
\documentclass[conference]{IEEEtran}
% Some Computer Society conferences also require the compsoc mode option,
% but others use the standard conference format.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

\usepackage{changepage}

@@

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  \usepackage[pdftex]{graphicx}
  \usepackage{epstopdf}
  % declare the path(s) where your graphic files are
 \graphicspath{{diagrams/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  \DeclareGraphicsExtensions{.pdf,.jpeg,.png,.eps}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  \graphicspath{{diagrams/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation
% can be obtained at:
% http://www.ctan.org/pkg/graphicx
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found at:
% http://www.ctan.org/pkg/epslatex
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
What is a developer's contribution to a repository? By using only commit counts and number of lines changed, existing tools that analyze code repositories fall short on showing the tangible contributions made by each developer. Individual commits provide the changed files, code diffs and a message. However, when many commits are viewed as a group, the details are lost. Currently there is a lack of tools for tracking the changes in a repository beyond simply looking at the number of commits of lines changed. The commit information can be misleading since lines of code gives no indication of what was actually being worked on without careful examination of the changed code. Providing a more comprehensive view of this information could provide deeper insights into how software projects evolve since changes to design and features are not clearly visible.

We present a novel method for visualizing Java projects. Instead of simply counting line changes in a commit we extract detailed type information over time by using the differences between Abstract Syntax Trees (ASTs). We are then able to track the additions and deletions of declarations and invocations for each type.

Using this tool we examine specific cases in well known repositories where our tool reveals interesting and useful information. We then compare type coverage information from the AST compared to file coverage to determine if unique information is provided by type information. 
\end{abstract}

@@

Source code is often represented as text in a source code file. This text is then lexed and parsed by a compiler or interpreter with the intent of eventually executing the instructions within the source code file. While programmers may type in characters, their intention is at a much higher level and more abstract. Their intent is written in tokens that represent statements, identifiers and types that are used to solve their problem. Programmers solve their problem in a way that is readable to other developers and executable by a computer. The programmer rarely operates at the character level, or even the line level, so why should our analyses of source code operate at granularities unfamiliar to developers?

The history and evolution of a software project is often tracked by version control systems such as Git in a programming language agnostic way. Updates to source code, documentation, and tests is recorded in the form of commits, which show the differences in lines between each subsequent revision of the software. Version control systems become vital for large projects since individual changes would be impossible to track manually. Large projects have thousands of individual changes made by many different developers. With more changes to peruse, it becomes more difficult to see how a project is evolving over time since there is so much information to look through. Most hosting services for Git repositories provide some basic tools for analysing projects over time~\cite{github-graphs,bitbucket-graphs}; but, as we will show, these tools often lack the kind of detail to gather insightful analysis. While commits may store individual changes to lines of code, they represent the change in \emph{structure} of the software.  By visualizing lines changed, or files changed, the current state-of-the-art for software project visualization does not capture the \emph{meaning} behind each commit; lines of code alone does not describe how a repository changed at a particular point in time. Thus, information regarding contributions from developers is often misleading since the number of commits and lines of code changed does not necessarily reflect meaningful changes to a repository~\cite{robles2014}.

In this paper, we propose a method that analyses the changes in type usage and method invocations over time, which we claim offers more insight than line-based, file-based, and commit-based methods. Our method is novel because none of the state-of-the-art tools has considered analyzing types changes over time or doing AST diffs.

In Java, a type can be a primitive type or an object reference. Primitive types are those that are built into the language such as an int or char. Object references point to instances of classes in memory which are defined in the project itself or through an included library. The types in a project represent the functionality available for a developer to work with. A large or complex project may have many different types, whereas a small project might only use a few. Different parts of a program use different types and external dependencies. When writing or contributing to a project, the author’s choice of which part of the program to edit will in turn affect what types they end up working with.

@@

Type information is useful for seeing what libraries and functions are being used.  When a part of code becomes deprecated, continuing to use it may lead to design or security flaws. Viewing the type information of a change can show when deprecated classes are added or removed from a project as well as how many remain. In a more general sense we can see when types are being added or removed from the project as well as what types are heavily used throughout the project.

@@

Dyer \textit{et al.} \cite{Dyer:2014:MBA:2568225.2568295} mined AST nodes to study the use of new Java language features over time. They found that all new Java features do get used, but not nearly as often as they could have been. New features varied greatly in popularity and adoption rate. Developers would also modify old code to use new features as they were released. The authors also found the most popular features and how teams adopted new features. They did not check to see how much a developer used a feature but instead only checked to see if they used it at all.

@@

Wagstrom \textit{et al.} \cite{Patrick:Wagstrom:2012} explored the roles that developers take in a networked, social development environments like GitHub. These role were defined as their level of contribution as well as the types of issues they dealt with. Developers will take on multiple roles in a project and will sometimes fulfill the same role across different projects.

@@

Wu \textit{et al.} \cite{wu2004a} explore punctuated changes to software repositories with a tool that generates an Evolution Spectrograph. Each file has a representative row in the chart, with each cell in the row representing a section of time. This diagram would show changes to incoming and outgoing dependencies by darkening the cell where the change happened and lightening it over time if no changes were made.

In another paper, Wu \textit{et al.} \cite{wu2004} the Evolutionary Spectrogram is used but with greater control of granularity for files and software subsystems such as directory structure. Spectrograms are also used to display the cardinality of a developer’s commits. The cardinality of a commit is the number of files or subsystems it affects in a system. All of the developers changes together show how many changes affect broad sections of the software.

Voinea \textit{et al.} \cite{voinea2006} visualizes CVS repositories using CVSgrab and adding their own enhancements to show changes to files over time. Files can be sorted by creation time, activity, or alphabetically and colour coded based on author, file size, file type or by search strings.

Kim \textit{et al.} \cite{Kim:2009:DRS:1555001.1555046} built a tool called Logical Structural Diff (LSdiff) which examines code changes and produces logical rules to represent them. Changes which do not match the rules are then pointed out to the developer. They also formed a focus group with software engineers who found that LSdiff complements existing tools well and showed promise for discovering inconsistent changes.

The GitHub tools have the strength that they do not rely on what type of code is being committed. It does not matter what language or errors exist in the commit GitHub will simply spit out the number of line additions and deletions. The issue with this type of analysis is that it does not give a clear breakdown of is being changed only that changes have occurred.

@@

\begin{figure}[!h]
\centering
\includegraphics[height=2.5in]{commits_view}
\caption{The commits increase in time from bottom up. Therefore, B and C are compared against A and E is compared with D. Since D is a merge commit it is ignored as its changes are recorded in C and B.}
\end{figure}

In order to gather all the changes for a given repository over time we generated ASTs for each commit and compared the difference. We did this by running ‘git log’ over the git repository and listing all the commit ids in increasing time. Each commit and its parent were then compared. If the commit had no parent then every file in the commit was treated as an addition. Since we compared each commit to its parent we were able to account for branching that happened in the master branch of the repository. Also we ignored commits that were merges as all changes in the merge should already be accounted for in previous commits.

\begin{figure}[!h]
\centering
\includegraphics[width=\columnwidth]{context}
\caption{}
\end{figure}

\section{The Tool}
For convenience we provide a variety of views to help organize the data. These are to provide a finer level of detail. The main view shows a stacked bar graph over time for all types. Where the blue and orange portions account for the number of additions and deletions respectively. The height of the bar shows the total number of changes in that time period with respect to the maximum number of changes in a single period.  The total size of the bar shows the total number of changes. This is calculated by multiplying the bar height by the number of changes during that time period divided by the max number of changes for this type. The width of each bar spans over a given time period and where the time period can be changed to span a hour, day, week, or month. The types are sorted from most to least used.

If a user mouses over a specific bar they are given some simple stats for the given type and time period. This includes the types full name, number of commits, number of authors, number of additions, number of deletions, and period start and end date. If the user then clicks on the bar they are given more detailed stats where we list the authors and commits.

@@

In the following sections, we present {N} case studies demonstrating the extra insight that TypeV method makes available.

@@

We chose Antlr4 because it is a substantial Java project with a commit history spanning several years. We can also see distinct changes throughout its history.

For example, in December of 2011 we can see that the ParserATNSimulator class has a large amount of additions and deletions in roughly equal proportion (Figure~\ref{fig:parser1}. This suggests that ParserATNSimulator is not being removed from the project but instead there is some refactoring being done. If we take a closer look at the commit messages we can see that they mention refactoring and reorganization of code (Figure~\ref{fig:parser2}). Zooming in on that date,  shows that during that month they were finishing work on a new version of ParserATNSimulator.

@@

\begin{table}[!t]
\renewcommand{\arraystretch}{1.3}
\caption{Comparision of Type Coverage and File Coverage in Projects}
\label{tab:summary}
\centering
\begin{tabular}{c|ccc}
\hline
\bfseries Project & \bfseries Wilcoxon & \bfseries Pearson (lin) cor & \bfseries Spearman (rank) cor \\
\hline
antlr4 & p$\ll$0.01 & 0.943 & 0.907\\
bookkeeper & 0.313 & 0.984 & 0.972\\
curator & 0.007 & 0.975 & 0.980\\
tika & 0.653 & 0.955 & 0.998\\
\hline
\end{tabular}
\end{table}

To determine if type coverage as returned by TypeV gives different information than file coverage alone, we computed the following. For four projects (Antlr, Apache Bookkeeper, Apache Curator, Apache Tika), we counted the file coverage (cumulative amount of files touched per author), as well as type coverage (cumulative amount of types touched per author) on a commit-by-commit basis. That is, each datum is represents a commit with two dimensions: file coverage and type coverage. Data is sorted in ascending order of commit date. Table~\ref{tab:summary} lists a summary of the data

@@

\subsection{Threats to Validity}

Authors can pull other projects into the repository and these additions will count towards the author's changes.

We do not look at other branches besides the master; therefore, we might miss some of the work done by an author.

We ignore the merge commits so we do not double count changes; however, if there is a merge conflict and the developer makes changes then these changes will not appear in TypeV. We make the assumptions that if this case happens then the changes are minor.

\subsection{Future work}

Currently for generic types we treat every instance of a generic as a different type. So for example if a user declares a list of strings and a list of integers these will appear as two different types. However, if they declare multiple instances of a list of strings then these instances will count toward the same type. In the future we would nice to come up with a way to better handle generic types.

Currently we are only looking at declarations and invocations, but it would be nice to also view what classes these changes were made to.

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\nocite{*}
\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{bibliography,autogen}
